---
title: "Synthetic Controls and Their Applications"
description: |
  Creating a comparison group when none is easily available
author:
  - name: Adam Peterson
    url: https://apetersonsite.com/
date: 03-15-2024
output: distill::distill_article
citation_url: https://xstreetvalidated.com
creative_commons: CC BY
categories:
  - Technical
draft: true 
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
#library(sf)
```

I've been wanting to do a write up on a series of papers that explore and 
discuss a relatively new and novel construct in the statistical literature:
synthetic controls. The first, as far as I can tell, paper that 
formalized this idea was a work by 
[Alberto Abadie](https://economics.mit.edu/people/faculty/alberto-abadie),
titled "The Economic Costs of Conflict: A Case Study of the Basque Country" 
[@abadie2003economic]. In this post I want to break down the ideas presented
in this paper, identify the key principles at work and explore how these same
ideas could be applied to other statistical methods - which we'll explore later
- as well as studies in the built environment.


## Background


```{r, fig.width = 4, fig.height = 2, fig.cap="By TUBS - This vector image includes elements that have been taken or adapted from this file:, CC BY-SA 3.0, https://commons.wikimedia.org/w/index.php?curid=14520652"}
## TODO(apeterson91): Consider replacing this image with a plot.
knitr::include_graphics("Pais_Vasco_in_Spain_(including_Canarias).svg")
```

## Method

## Application 

## Extensions in Methods and Application