---
title: "Bowling Alone: The Collapse and Revival of American Community Revised and Updated"
description: |
  A review of the seminal text on social capital
author:
  - name: Adam Peterson
    url: https://apetersonsite.com/
date: 12-24-2023
output: distill::distill_article
citation_url: https://xstreetvalidated.com
creative_commons: CC BY
categories:
  - Book Review
  - Social Capital
draft: true 
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(haven)
library(sf)
library(srvyr)
library(tidyverse)
library(tidycensus)
```

```{r plot_theme, include=FALSE}
theme_set(theme_bw() +
  theme(
    strip.background = element_blank(),
    text = element_text(size = 22)
  ))
```

> “In small ways like this – and in larger ways, too – we Americans need to
reconnect with one another. That is the simple argument of this book.”

## Introduction

One of the very first themes discussed in the motivation for starting this blog
was the notion of “social cohesion.” In a small post filled with imprecise
terms, social cohesion stood out as perhaps the most imprecise. In an attempt
to pin it down, I referred to ideas like crime, civil unrest, and a
“feeling of belonging.” In Bowling Alone: The Collapse and Revival of American 
Community, Robert Putnam makes one of the best efforts I’ve seen to speak more 
precisely and quantitatively about this nuanced and subtle 
topic[@putnam2000bowling]. He chooses a
better term, too: social capital. 


Inspired by his previous research exploring the relationship between civic
engagement and healthy democracy, Putnam’s groundwork for writing Bowling Alone 
was a 1995 article that detailed Americans’ participation in various civic 
groups or activities during the 20th century[@putnam1995tuning]. Hypothesizing 
that a decline in these forms of participation would correspond to a 
proportionate increase in social–political problems, Putnam set out to document 
the state of US social capital in the decades between 1960 and 2000.

Putnam published the results of his work in the first edition of Bowling Alone 
in 2000. Two decades later, he published a revised and updated edition of 
Bowling Alone alongside a new book, The Upswing. This new entry broadened the 
scope of the story, examining changes in social capital from the late 1800s to 
2020. The story he tells in the latter book is one that might sound increasingly
familiar - in a time of growing inequality, unhealthy civil dialogue and severe 
political polarization, it seems like the only thing Americans can agree on is 
that they disagree about everything. Putnam points out this was as true at the 
turn of the 20th century as it has in the 21st century. What happens in between 
is a story of America coming together in the face of two World Wars and then 
subsequently falling apart again amidst the rise of a new generation embroiled 
in social change. Putnam argues in both books that we should – must – come 
together again, as the challenges we face in 2020 are even greater than those 
faced in the 20th century.

I plan to focus my commentary predominantly on the material in Bowling Alone, 
but I’ll do my best to ground it in a modern context. In doing so, I hope to 
highlight both the relationship between the built environment and the various 
social measures that are our best proxy for measuring social capital and the 
importance of social capital itself. While “social capital” broadly refers to 
the notions of security and belonging that I mentioned previously, at its heart,
it's a measure of the strength of human relationships. As most individuals 
report, their relationships are the most important and fulfilling part of their 
lives. I think it's important to take note of how our environments can be more 
or less supportive of their growth and stability.

data available [here](http://bowlingalone.com/?page_id=7)

```{r}
df <- read_tsv("data/bowling_alone_data/DDB master 34 (1975-1998).dat") %>%
  mutate(
    State = factor(STATE, labels = c(
      "Connecicut", "Maine",
      "Massachusetts", "NewHampshire",
      "Rhode Island", "Vermont",
      "New Jersey", "New York",
      "Pennsylvania", "Illinois",
      "Indiana", "Michigan",
      "Ohio", "Wisconsin",
      "Iowa", "Kansas",
      "Minnesota", "Missouri",
      "Nebraska", "North Dakota",
      "South Dakota", "Delaware",
      "District of Columbia",
      "Florida", "Georgia",
      "Maryland", "North Carolina",
      "South Carolina", "Virginia",
      "West Virginia",
      "Alabama", "Kentucky",
      "Mississippi", "Tennessee",
      "Arkansas", "Louisiana",
      "Oklahoma", "Texas",
      "Arizona", "Colorado",
      "Idaho", "Montana",
      "Nevada", "New Mexico",
      "Utah", "Wyoming",
      "California", "Oregon",
      "Washington"
    )),
    Density = factor(POPDEN1, labels = c("Urban", "Rural", "Urban", "SMA", "Urban", "Rural", "SMA", "Urban", "Rural", "SMA", "Urban", "Rural"))
  )
```


```{r}
df %>%
  select(YEAR, WEIGHT, ENTHOME, DINPARTY, PICNIC, FAMDIN) %>%
  filter(!is.na(ENTHOME), !is.na(DINPARTY), !is.na(FAMDIN), !is.na(PICNIC)) %>%
  mutate(
    `entertained` = (ENTHOME > 1),
    `dinner party` = DINPARTY > 1,
    `family dinner` = (FAMDIN >= 4),
    picnic = (PICNIC > 1)
  ) %>%
  gather(entertained, `dinner party`, `picnic`, `family dinner`,
    key = "social_capital", value = "measure"
  ) %>%
  as_survey_design(weights = WEIGHT) %>%
  group_by(YEAR, social_capital) %>%
  summarize(avg = survey_mean(measure, vartype = "ci")) %>%
  ggplot(aes(x = YEAR, y = avg, color = social_capital)) +
  geom_pointrange(aes(ymin = avg_low, ymax = avg_upp)) +
  geom_line() +
  theme(legend.title = element_blank(), legend.position = "top") +
  scale_color_brewer(palette = "Dark2") +
  ylab("% Of U.S. Population") +
  xlab("") +
  scale_y_continuous(labels = scales::percent) +
  ggtitle("Decreases in informal social connections 1975-1998") +
  labs(caption = "Point indicates mean, line represents 95% confidence interval. \n Data from DDB survey data")
```


## The Time It Takes to Get There: A Causal Account for the Decrease in Social Capital


```{r, echo = FALSE, eval=FALSE}
## commute time doesn't change over years observed
df %>%
  select(YEAR, WEIGHT, COMMUTIM) %>%
  filter_all(function(x) !is.na(x)) %>%
  as_survey_design(weights = WEIGHT) %>%
  group_by(YEAR) %>%
  summarize(min = survey_quantile(COMMUTIM, .05, vartype = "ci")) %>%
  ggplot(aes(x = YEAR, y = min_05)) +
  geom_pointrange(aes(ymin = min_05_low, ymax = min_05_high))
```


```{r}
## TV associated with antisocial behavior
tvdf <- df %>%
  select(
    YEAR, WEIGHT, TVENTER, FINGER, DINPARTY, CLUBMEET,
    PICNIC, ENTHOME, GREETING, POLITICS, BLOOD, LDPHOCL,
    SNTEMAIL
  ) %>%
  mutate(
    tv_viewer = if_else(TVENTER >= 4, "TV is my Primary Form of Entertainment",
      "TV is *not* my Primary Form of Entertainment"
    ),
    dinner_party = if_else(DINPARTY > 1, 1, 0),
    club_meetings = if_else(CLUBMEET > 1, 1, 0),
    picnic = if_else(PICNIC > 1, 1, 0),
    entertain = if_else(ENTHOME > 1, 1, 0),
    greeting_card = if_else(GREETING > 1, 1, 0),
    # interested in politics = 1
    politics = if_else(POLITICS >= 4, 1, 0),
    ## donated blood
    blood = if_else(BLOOD > 0, 1, 0),
    ## made long distance call
    long_distance = if_else(LDPHOCL > 1, 1, 0),
    ## sent an email
    email = if_else(SNTEMAIL == 1, 1, 0),
    ## gave someone the middle finger on the road
    finger = if_else(FINGER > 1, 1, 0)
  ) %>%
  gather(dinner_party, club_meetings, picnic, entertain,
    greeting_card, politics, blood, long_distance, email, finger,
    key = "social_capital", value = "measure"
  ) %>%
  select(YEAR, WEIGHT, tv_viewer, social_capital, measure) %>%
  filter(!is.na(measure))
```

```{r}
tvdf %>%
  filter(
    !is.na(measure), !is.na(tv_viewer),
    social_capital %in% c(
      "club_meetings", "dinner_party", "entertain",
      "greeting_card", "picnic", "politics"
    )
  ) %>%
  mutate(social_capital = case_when(
    social_capital == "politics" ~ "Interested in Politics",
    social_capital == "picnic" ~ "Attended a Picnic",
    social_capital == "entertain" ~ "Entertained Friends",
    social_capital == "greeting_card" ~ "Sent a Greeting Card",
    social_capital == "dinner_party" ~ "Hosted a Dinnery Party",
    social_capital == "club_meetings" ~ "Attended a Club Meeting"
  )) %>%
  as_survey_design(weights = WEIGHT) %>%
  group_by(YEAR, social_capital, tv_viewer) %>%
  summarize(avg = survey_mean(measure, vartype = "ci", proportion = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = YEAR, y = avg, color = tv_viewer)) +
  geom_pointrange(aes(ymin = avg_low, ymax = avg_upp)) +
  geom_line() +
  theme(legend.title = element_blank(), legend.position = "top") +
  scale_color_brewer(palette = "Dark2") +
  ylab("% Of U.S. Population") +
  xlab("") +
  facet_wrap(~social_capital) +
  scale_y_continuous(labels = scales::percent) +
  ggtitle("TV Association with Less Social Behavior.",
    subtitle = "Respondents Self Reported Activities in Last Twelve Months."
  ) +
  labs(caption = stringr::str_c(
    "Point indicates mean, line represents 95%",
    "confidence interval. \n Data from DDB survey data"
  ))
```

```{r}
volunteer_tv_des <- df %>%
  select(
    YEAR, WEIGHT, TVENTER, VOLUNT, INCOME, SEX, RACE,
    RESAGERG, EDUC,
  ) %>%
  mutate(
    tv_viewer = factor(if_else(TVENTER >= 4, "TV is my Primary Form of Entertainment",
      "TV is *not* my Primary Form of Entertainment"
    )),
    ## TODO add labels
    volunteered = if_else(VOLUNT > 1, 1L, 0L),
    income = factor(INCOME),
    sex = factor(SEX),
    age = factor(RESAGERG),
    edu = factor(EDUC),
    race = factor(RACE),
  ) %>%
  filter_all(function(x) !is.na(x)) %>%
  as_survey_design(weights = WEIGHT)
```

```{r}
fit <- survey::svyglm(volunteered ~ tv_viewer + YEAR + race + income +
  sex + age + edu,
design = volunteer_tv_des,
family = binomial()
)
```

```{r}
broom::tidy(fit)
```

## Other Resources on Bowling Alone
