---
title: "Cleanup vs. Crime: Causal Inference"
description: |
  What kinds of causal claims are justified from an empirical analysis?
author:
  - name: Adam Peterson
    url: https://apetersonsite.org
date: 11-28-2021
creative_commons: CC BY
output:
  distill::distill_article:
    self_contained: false
bibliography: cleanup.bib
categories:
  - Technical
  - Cleanup vs. Crime
  - Social Capital
draft: false 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse) 
library(sf)
library(patchwork)
theme_set(theme_bw() + theme(text = element_text(size=22)))
```

```{r}
#OPA property assessments - includes a Category Code where 6 = vacant land - this is available online in several forms at https://www.opendataphilly.org/dataset/opa-property-assessments
```

After we've calculated our buffer counts and fit our fancy mixed effects 
regression model what kind of claims can we make about the results?

In the [paper](https://ajph.aphapublications.org/doi/full/10.2105/AJPH.2018.304752)
we've been studying, the author's estimate a decrease in the number of crimes 
around vacant lots randomly assigned the greening or mowing interventions, 
stating in their abstract:

>Conclusion: Remediating vacant land with inexpensive, scalable methods, including greening or minimal mowing and trash cleanup, significantly reduced shootings that
>result in serious injury or death.

>Public Health Implications: Cities should experiment with place-based interventions
>to develop effective firearm violence–reduction strategies.

Both these and other claims the authors might make require varying
level of assumptions in order to be valid.  In this post, we'll walk through 
these claims step-by-step, identifying the tools the author's did or did not 
use to try to fulfill each assumption.

# Descriptive Statistics vs. Causal Inference 

Let's start with the easiest, most defensible claim the authors can make 
after finishing their analysis. I'll limit these statements to just the greening intervention since that had the more robust results across the models.

"The greening intervention was **associated** with a decrease in crimes
in the area proximal to the vacant lots in our **sample**."

There are two keywords that I've bolded in the above quote. The first ---
associated --- is important because it implies that there is not necessarily a 
causal relationship between the greening intervention nearby crime. 
The second --- sample --- implies that the association may be limited to the 
sample and not generalizable to a larger population of vacant lots and 
shooting crimes generically found in the world at large. 

These claims fall under the category of what statisticians might 
colloquially call "descriptive", in that they are absolutely true in describing
the sample as observed. Scientific research generally aims to establish more 
general truths that describe the causal relationships that determine the 
data we observe. In order to make the jump from describing our sample to 
describing the world, there are several assumptions we have to make, critical
to our ability to draw a **causal conclusion** from our descriptive analysis.

## Assumptions Assumptions 

Here's a stronger claim than the previous. 

"The decrease in crime observed was *caused* by the greening intervention in 
our sample."

In order to make the jump in claim from association, or correlation, 
to causation the author's need to defend their ability to have met the 
identifiability assumptions for a causal effect. I've discussed these ideas in greater depth 
[here](https://www.apetersonsite.org/post/causal_inference_preface/)
and [here](https://www.xstreetvalidated.com/post/causal-inference-in-the-built-environment)
but for now we'll focus on (1) the author's strategy in using their second buffer model
to check for spillover effects and how this relates to the notion of 
of interference, as well as (2) how the author's use of randomization **should** 
provide them with some justification for arguing that the intervention assignment is independent of other potential confounding factors.

### Spillover and Interference 

If a cluster of vacant lots were assigned to the greening intervention, 
how far would the decrease in crime associated with that intervention extend out
in space? Is it possible that the effect could impinge upon another vacant lot? 

<aside>
Here we're considering the *spatial scale* of the intervention. 
Just as potentially important is the *temporal scale*, or that time at 
which the intervention has its greatest impact. See [@peterson2021spatial] for 
more on this idea.
</aside>

```{r bufferinterferevis,fig.width = 14,fig.height=8,fig.cap="Illustration of buffer zones around hypothetical vacant lot. First buffer in green is used to detect proximal increase/decrease in crime, while the second is used to check for spillover effects by considering the non-overlapping area between the two circles.", cache = TRUE}
uk <- read_sf("~/Documents/CityData/Philly/geo-data/Neighborhoods_Philadelphia/Neighborhoods_Philadelphia.shp") %>%
  filter(MAPNAME == "Upper Kensington")  %>% 
  st_transform(4326)

shootings <- read_csv("~/Documents/CityData/Philly/Crime/shootings_2015_2020.csv") %>% 
  filter(!is.na(point_x),!is.na(point_y)) %>% 
  filter(point_x > -80,
         year==2015,
         officer_involved == "N") %>% 
  st_as_sf(coords = c("lng","lat")) %>% 
  st_set_crs(4326) %>% 
  select(year) %>% 
  st_filter(uk)

streets <- read_sf("~/Documents/CityData/Philly/CompleteStreets-shp/CompleteStreets.shp") %>% 
  st_filter(uk)

hypothetical_lot <- st_centroid(uk)

buffer_one <- st_buffer(hypothetical_lot,dist = units::as_units(300,"m"))
buffer_two <- st_buffer(hypothetical_lot,dist = units::as_units(600,"m"))

p1 <- uk %>% 
  st_join(streets,join=st_contains) %>% 
  ggplot() + 
  geom_sf(linetype=2) + 
  geom_sf(data=streets) + 
  geom_sf(data=shootings,color='red') + 
  geom_sf(data=hypothetical_lot,color='blue',shape='*',size=13) + 
  geom_sf(data=buffer_one,color='darkgreen',alpha=0,size=1) + 
  geom_sf(data=buffer_two,color='#502989',alpha=0,size=2) + 
  theme_void() + 
  theme(text = element_text(size=22)) + 
  labs(caption = "Data sourced from OpenDataPhilly.org")

p1
```

Figure \@ref(fig:bufferinterferevis) illustrates the method the author's used
to check for "spill-over". By looking both 300 and 600 meters around the vacant
lots they sought to establish that the interventions did not simply move the 
shootings from closely to slightly further away, but actually did diminish the 
overall number of shootings. However! What the author's failed to acknowledge
and comment on in their paper, was the impact that this extended spatial 
impact --- when one vacant lot's effect may impinge upon another --- may have 
upon their estimates of the intervention effect, as it demonstrates the capacity
for these vacant lot interventions to **interfere** with one another. Recall 
that the author's clustered vacant lots in order to try and limit the impact from 
spatial proximity, but this extended spatial effect shows the effect 
may go farther than previously anticipated. We'll use 
this as a point of discussion in a future blog-post but for now we'll simply
consider it a question this result leaves us asking.

```{r interfere, cache = TRUE, fig.width = 14, fig.width = 8,fig.cap="Visualization of hypothetical vacant lots proximal to one another whose effects may interfere with one another despite clustering."}
pts <- st_sample(buffer_one, 2)
pts <- st_sf(pts) %>% 
  mutate(id = 1:n()) %>% 
  rename(geometry=pts)
pts <- hypothetical_lot %>% select(geometry) %>% mutate(id=0) %>% rbind(pts)
gd <- expand.grid(point_one = 1:nrow(pts), point_two = 1:nrow(pts)) %>% 
  filter(point_one != point_two)

ptpltdf <- pts %>% 
  as_tibble() %>% 
  left_join(pts %>% 
              as_tibble(), 
            by=character()) %>% 
  filter(id.x!=id.y) %>% 
  mutate(geometry.z = st_cast(st_union(geometry.x,geometry.y),"LINESTRING")) %>% 
  select(geometry.z) %>% mutate(id = 1:n()) %>% 
  st_as_sf()

uk %>% 
  ggplot() + 
  geom_sf() + 
  geom_sf(data=streets,alpha=.3) + 
  geom_sf(data=hypothetical_lot,color='blue',shape='*',size=13) + 
  geom_sf(data=pts,color='red',shape = '*',size=12) + 
  geom_sf(data=ptpltdf,color='black',linetype=2) + 
  theme_void()
```

### Repeat Randomization

Another tool the author's used was randomization: "A repeat randomization
procedure ensured that the clusters were balanced on potential confounding 
variables including the total area and mean separating distance of the vacant
lots, the resident population, and the number of Summary Reporting System
Part I serious violent and property crimes, such as robbery and burglary."
[@moyer2019effect].

The authors use randomization to try and create *exchangeability* between the
different intervention groups. That is that there is just as much likelihood
of a crime occurring in the area proximal to a control vacant lot as it would
in an area proximal to a vacant lot receiving one of the two interventions.

There are two things to note here, (i) the author's are doing everything
they can to try and ensure that these different vacant lot clusters are 
comparable across the different intervention groups and (ii), they 
seem to have trouble accomplishing this goal. 

As they state in their results, "The greened lots tended to have a 
slightly higher average shooting rate in the baseline period (pre), which we 
accounted for in the regression estimates".  This *imbalance*^[See Chapter 20 
Section 4 in [@gelman2020regression] for more on this topic.] between
the different intervention groups, reflects a difficult truth in built 
environment research and the larger realm encompassing the social sciences
more generally: it is really hard to make things that are fundamentally
different comparable for the sake of analysis. This may be why the authors 
chose to conduct their analysis using the difference-in-difference 
estimators, as this is not the strongest, or "best" contrast that 
a scientist typically hopes to make in establishing a causal effect.

## Treatment Effects

Let's continue with our hierarchy of claims. 

"Applying greening interventions **caused** there to be a reduction in crime 
around vacant lots **greater** than one would observe otherwise."

This is a very specific kind of claim that may be more familiar when considered
in the context of a pharmaceutical clinical trial. If we were to randomly assign
a treatment and placebo to a group of individuals who had some disease,
we would be most interested in the difference of outcomes between the individuals
who had the treatment and those who had the control. In statistics and causal 
inference this is called the Average Treatment Effect (ATE), and 
is the most frequent causal estimand of interest. If $A_i=1$ 
indicates the $i$th subject, $(i=1,...n)$, in a  clinical trial received the 
treatment and $A_{i'}=0$ indicates the $i'$th subject received the placebo and 
we were measuring some outcome $Y^{(t)}_i$ at time points $t = 0, 1$, then we 
would define the ATE and estimated ATE as follows:

\begin{equation}
\begin{aligned}
ATE &= E[Y^{1}|A=1] - E[Y^1|A=0] \\
\hat{ATE} &= \frac{1}{n}\sum_{i=1}^{n} Y_i^1A_i - \frac{1}{n} \sum_{i=1}^{n}Y_{i}^1(1-A_i)
\end{aligned}
\end{equation}

Another causal estimand is the Average Treatment Effect on the Treated or 
ATET. This estimand is represented by the following expression, note the 
difference in the time points and treatment variables :
$$
\begin{aligned}
ATET &= E[Y^{1}|A=1] - E[Y^{0}|A=1].\\
\hat{ATET} &= \frac{1}{n}\sum_{i=1}^{n} Y_i^1A_i - Y_i^0A_i
\end{aligned}
$$

Conceptually the regression models the author's used, discussed in the previous 
post, estimates this latter (ATET) entity, but in a slightly more complicated 
way.
To make things simpler, this idea can be represented as the difference 
between the two points connected by the line in Figure \@ref(fig:ATEvis) below.

```{r ATEvis, cache = TRUE, fig.width=14,fig.height=8,fig.cap="Visualization of Average Treatment Effect (ATE) and Average Treatment Effect on the Treated(ATET) below using the crime density estimates from [@moyer2019effect]. Solid line is above the two points involved in the ATET and the dotted line is above the two points involved in calculating the ATE.", preview = TRUE}
pltdf <- tibble(Group = c("Greening","Mowing","Control",
                          "Greening","Mowing","Control"),
                Crime = c(1.25,1.2,1.09,.98,.91,.89),
                Period = factor(c(rep("Pre",3), rep("Post",3)), 
                                levels = c("Pre","Post")))  

p1 <- pltdf %>% 
  ggplot(aes(x=Period,y=Crime,color=Group,group=Group)) + 
  geom_point() + 
  geom_line() + 
  xlab("") + 
  labs(y = bquote('Avg # Crimes /' ~ km^2)) + 
  theme(legend.position = "top",
        legend.title = element_blank()) + 
  geom_segment(x = 2.05, xend = 2.05, y = 1.01, yend = 1.25,
               color='black') +
  geom_segment(x = 2.05, xend = 2.05, y = 1.0, yend = .89,
               color = 'black', linetype = 2) + 
  annotate("text", x = 2.15, y = 1.05, label = "ATET", size = 5) + 
  annotate("text", x = 2.15, y = .95, 
           label = "ATE", size = 5) + 
  ylim(0,1.5)

p1 
```


Bringing these ideas back to the claim at hand, the author's did *not*
aim to justify this statement. By using a difference-in-difference 
estimator they estimated the ATET, which was their "take-home" result.


While this is still an interesting result, it does make a reader like me wonder 
why they went to the effort of randomizing interventions as they did, if they 
didn't even use their control vacant lots in the contrast estimate of the 
intervention effect(s). My guess is that the result was not significantly 
different and they didn't want this result complicating their story. 

However, before we get too dramatic, it is important to realize that the authors
do have some justification for focusing on the ATET as they did. 
As mentioned above, the pre-intervention crime measures for the three different 
vacant lot groups were all different, suggesting these may not have been as 
comparable as they should have been from randomization. 
Given that, it seems that one of the best and most conservative things the 
authors could do, would be to focus on the ATET as they did. Still, it 
is not the best one might hope for when looking to advance the cause 
of "place-based interventions" as I and the authors both hope to do.

<aside>
Note that under an additional [assumption](https://en.wikipedia.org/wiki/Difference_in_differences#Assumptions),
termed the parallel trend assumption, the ATET is equal to the ATE. This 
assumption is often made in natural experiments, where one cannot randomize 
a treatment and control group. However, it is not likely relevant here,
as the author's could explicitly compute the ATE.
</aside>


## Generalizability

"Greening vacant lots will reduce in crime in the surrounding area *generally*
and not just in Philadelphia"

Separate from the ideas of ATE's, ATET's and randomization is the question of 
generalizability. How much can we hope that the results found here will be
reproduced in other cities around the globe? This likely depends on the 
mechanism by which violent crimes manifest, as well as how greening 
or mowing vacant lots plays into this mechanism.  The authors speak to this 
idea in part in their discussion, and we'll discuss this idea further in the 
subsequent post discussing the substantive reasoning at play in this study. 
For now it might suffice to say that the more similar a city is to Philadelphia
the more likely one would be to feel justified in expecting these results to
apply and vice versa. 

## Recap and Conclusion

Let's review the different claims made and how they each measured up in 
our analysis.

1.  "The greening intervention was **associated** with a decrease in crimes
in the area proximal to the vacant lots in our **sample**."
  

This claim has the strongest level of evidence for it. In fact I'd say it is 
trivially true, barring any serious data collection or study design errors.
  

2.  "The decrease in crime observed was *caused* by the greening intervention in 
our sample."

This claim is in line with the author’s conclusion in their abstract and 
is supported mainly by the ATET estimate.
However, there is still the potential for interference --- despite clustering --
to attenuate this estimate or reformulate how the effect manifests. 

3.  "Applying greening interventions **caused** there to be a reduction in crime 
around vacant lots **greater** than one would 
observe otherwise."

The author's can't defend this claim. While they tried to make their 
intervention and control groups as comparable as possible through randomization,
baseline crime measures between the groups were quite different, forcing the 
author's  to abandon any contrast estimate between groups and focus only 
on within group differences.

4. "Greening vacant lots will reduce in crime in the surrounding area *generally* 
not just in Philadelphia"

Yes and no. Cities are highly variable and not at all homogeneous. The more 
similar a city is to Philadelphia, the more likely the results may apply. 
However, this is true for every built environment or place-based intervention, 
and shouldn't be held against the author's good-faith effort here. Despite 
the fact that the author’s can’t completely defend this claim, I agree with the 
statement they make in their abstract, under “Public Health Implications”, 
that other cities should adapt the techniques and lessons tried here. Given 
The cost/benefit ratio, there certainly seems enough evidence to try *something*,
Given the potential payoff.

Overall, in my judgement, this study is still quite strong, with only aspects of 
interference related to point 2. and lack of comparability related to points 3 and 4. 
holding them back from making all the claims that one could hope for from 
this work. In the next and final post of this case study, we’ll look through the 
mechanisms by which the decrease in crime may have occurred.

## Acknowledgements {.appendix}

Thanks to [Robert Svoboda](https://www.linkedin.com/in/rcsvoboda/) and [Paul Haluszczak](https://www.linkedin.com/in/paul-haluszczak-96805737/) for reading 
and offering comments on this article.
