---
title: "Housing Economics"
description: |
   A Review of "The Economic Implications of Housing Supply"
author:
  - name: Adam Peterson
    url: https://apetersonsite.com/
date: 9-01-2025
draft: true
output: 
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
citation_url: https://xstreetvalidated.com
creative_commons: CC BY
bibliography: housing.bib
categories:
  - Urban Economics
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(haven)
library(tidyverse)
library(tidycensus)
theme_set(theme_bw() + 
          theme(text = element_text(size = 22),
                strip.background = element_blank()))
here::i_am("_posts/2025-08-09-housing-economics/housing-economics.Rmd")
data_path <- here::here("_posts/2025-08-09-housing-economics/data")
# Wharton Residential Land use index sourced from
# https://real-faculty.wharton.upenn.edu/gyourko/land-use-survey/
```

## A First Foray into Urban Economics 

Digging more deeply into the subject of urban economics, as I mentioned in my 
most recent 
[book review](https://xstreetvalidated.com/posts/2024-12-28-order-without-design/) 
of "Order Without Design", has been no small affair. The subject covers a wide
array of topics that are each of interest in their own right. The list includes 
["agglomeration economics"](https://en.wikipedia.org/wiki/Economies_of_agglomeration),
[transport economics](https://en.wikipedia.org/wiki/Transport_economics) and 
more. For my first post within this subject though, I wanted to learn more
about housing and housing markets.

My reasons for this are many:

1. I've put away money for several years now with the idea of it being used as a
down payment and have often wondered at how houses got *so* expensive. 

2. Related to the first point, this topic is increasingly relevant:

    * It featured [prominently](https://www.ap.org/news-highlights/elections/2024/housing-on-the-ballot-harris-trump-push-different-plans-for-tackling-housing-affordability-crisis/) in the 2024 US Presidential Campaign.
    * It continues to dominate City Council [legislation](https://denverite.com/2025/08/05/denver-parking-minimums-abolished/).
    * [Advocacy groups](https://yimbyaction.org) have formed around the issue, pushing for more or less
    supply.

3. Finally, I think there are some very interesting cultural and moral questions
that surround housing. Both in terms of how it represents a fundamental need
of all humans (shelter) as well as an aspiration --- the American dream
is often associated with owning a single family house. 


I'll leave these last questions for the end of the post. To begin with, I'll 
draw from two sources to form the foundation of our technical understanding of 
housing markets. The first, [@o1996urban], will be a rudimentary *theoretical* 
introduction to the topic, drawing on a basic understanding of Supply and Demand.
If you're not familiar with these topics, I'd suggest 
[reviewing](https://www.youtube.com/watch?v=PEMkfgrifDw&ab_channel=KhanAcademy) 
them.


For the majority of this post, we'll review a paper by 
[Ed Glaeser](https://share.google/Ucl8neKLs8m0593Fy) and 
[Joseph Gyourko](https://share.google/Y5KVNvH0ZzxluK19P) titled 
"The Economic Implications of Housing Supply" [@glaeser2018economic]. 
This paper provides a nice overview of how the housing market has functioned 
*empirically* and also provides an opportunity to review their study with more
updated data.

```{r}
#| label: thumbnail-fig
#| fig-cap: "The ol' Supply and Demand Curve"
line1 <- data.frame(x = c(1, 9), y = c(1, 9))
line2 <- data.frame(x = c(9, 1), y = c(1, 9))
line_intersection <- reconPlots::curve_intersect(line1, line2)
ggplot(mapping = aes(x = x, y = y)) +
  geom_line(data = line1, color = "red", size = 1) +
  geom_line(data = line2, color = "blue", size = 1) +
  geom_vline(xintercept = line_intersection$x, linetype = "dotted") +
  geom_hline(yintercept = line_intersection$y, linetype = "dotted") +
  theme_classic() +
  xlab("Quantity") +
  ylab("Price") +
  ggtitle("Classic Market Equilibrium")
```

## An Introduction to Housing Market Theory

O'Sullivan starts his chapter on Housing by noting that housing is a **durable** 
good. This means that, in contrast to the apple bought at the supermarket, a 
house doesn't "go bad" after its been used by a particular occupant or occupants
but can be resold later, especially if it has been maintained well. Secondly,
the housing market is heterogeneous; there are different types of houses in 
terms of age, size, layout, etc. that can be purchased.

All of this filters into what Sullivan argues is a "Filtering Model" for the
housing market. Whereas in a "classic market", like our apple example,
there is no interaction between the supply of a given product at a previous 
point in time and the present, housing supply has an explicit dependence on
previous instances of supply. Sullivan formulates the model like this:

$$
H_t = H_{t-1} \times (1-f) + \text{new},
$$

where $H_t$ represents the quantity of housing supplied at an arbitrary 
time point $t$, dependent on the quantity at a previous time point $t-1$ 
multiplied by some drop out rate $f$, plus the addition of newly constructed 
housing, "new".

```{r}
#| label: housing-model
#| fig-cap: "Housing Filtering Supply Model"
q <- c(0, 20, 40, 50)
p <- c(108, 115, 200, 250)
house_supply <- tibble(quantity = q, price = p, display = c("",letters[1:3]))

house_supply |> 
  ggplot(aes(x = q, y = p)) +
  geom_line() +
  geom_point() +
  xlab("Quantity Supplied") +
  ylab("Price / Unit") +
  geom_text(aes(label = display), position = position_nudge(y = 5)) +
  ggtitle("Filtering Model of Housing Supply")
```

## The Economic Implications of Housing Supply

```{r}
#| label: wharton-reg-index
land_reg <- read_dta(file.path(
  data_path,
  "WHARTON LAND REGULATION DATA_1_24_2008.dta"
)) |>
  mutate(year = 2008) |>
  select(year, msaname99, state, weight, WRLURI)
```

```{r}
#| label: american-housing-survey
#| eval: FALSE
ahs <- read_sas(file.path(data_path, "ahs2007n.sas7bdat")) |>
  select(CPRICE, LPRICE, contains("REPWGT"), REGION, METRO3, CMSA, SMSA) |>
  rename(
    CurrentPrice = "CPRICE",
    LandPrice = "LPRICE",
    CensusRegion = "REGION",
    Metro = "METRO3"
  )
```

## Acknowledgements {.appendix}
Thanks to [Paul Haluszczak](https://www.linkedin.com/in/expertly-paul) for
reading and offering comments on this post.
